---
title: "Does Age Matter in Hollywood Casting? What the Numbers Say"
subtitle: "Assignment 4 ETC5512"
author: "Parth Tendulkar"
format: html
editor: visual
---

```{r setup, include = TRUE, echo = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

### Research Question

<big> **Does the age and gender dynamics of lead actors influence the commercial success of movies?**</big>

Hollywood has historically exhibited a noticeable pattern in casting, where older male actors are frequently paired with significantly younger female counterparts—a trend far less common in reverse. This project aims to examine whether such casting decisions are financially beneficial or merely reflective of entrenched industry biases. To explore this, two datasets were utilized: one containing information on actor ages and gender, and another detailing movie budgets, revenues, and profits. Together, these datasets provided a suitable foundation for analyzing the financial implications of age and gender dynamics in casting.

### Data Being used

**This analysis draws on two publicly available datasets:**

-   [**Movie Revenue Dataset**](https://github.com/ntdoris/movie-revenue-analysis/blob/main/README.md): This dataset includes key financial indicators such as production budget, domestic and international gross earnings, total profit, return on investment (ROI), and film genre.
-   [**Hollywood Age Gap Dataset**](https://hollywoodagegap.com): This dataset captures detailed information on romantic leads in films, including actor names, their ages at the time of the film’s release, character genders, and the age differences between paired actors.

**These datasets are well-suited for the objectives of this study for the following reasons:**

-   They collectively address both financial performance and casting patterns based on age and gender.
-   Relevant variables (e.g., age, gender, profit, ROI) are clearly defined and readily available.
-   The datasets are cleanly structured and easy to process for exploratory and comparative analysis.
-   The Hollywood Age Gap dataset does not have a formal license but permits usage with proper attribution, which has been acknowledged.
-   Both datasets are publicly accessible and appropriate for the level of analysis required in this assignment.

### Downloading Data

**Steps to Download *Hollywood Age Gap* Dataset Data**

-   Navigate to the official website: [**www.hollywoodagegap.com**](https://hollywoodagegap.com)\
-   Click on the **"Download Data"** button located in the **Top Right Corner** of the homepage.
-   Save the downloaded **CSV file** into the `data` folder within the project directory to ensure proper access during analysis.

**Steps to Download *Movie Revenue Dataset* Dataset Data**

-   Go to the **movie-revenue-analysis** Github Repository:\
    visit <https://github.com/ntdoris/movie-revenue-analysis>
-   Locate the dataset:\
    In the repo, go to the `data` folder. and locate `final_dataset.csv`.
-   Download the file:
    -   Click on the dataset file.
    -   Click the **"Download raw file"** button.\
-   Save the downloaded **CSV file** into the `data` folder within the project directory to ensure proper access during analysis.

### Data Analysis

1.  **Load Libraries**

```{r Load Libraries}

library(tidyverse)
library(here)
library(readr)
library(ggplot2)
library(patchwork)
library(scales)

```

2.  **Read in the Data**

```{r Read Data}

age_gaps_raw <- read_csv(here("data/movies.csv"))

revenue_raw <- read_csv(here("data/final_dataset.csv"))

```

3.  **Cleaning Data**

```{r}

age_gaps_clean <- age_gaps_raw %>%
  mutate(join_title = str_to_lower(str_trim(`Movie Name`)))

revenue_clean <- revenue_raw %>%
  mutate(join_title = str_to_lower(str_trim(movie)))

```

4.  **join datasets**

```{r Join Datasets}

joined_data <- age_gaps_clean %>%
  left_join(revenue_clean, by = c("join_title", 
                                  "Release Year" = "year"))

#filtering out non matches
joined_data <- joined_data %>%
  filter(!is.na(production_budget))

```

5.  **create new variable**

```{r new_variables}

joined_data <- joined_data %>%
  mutate(
    `Older Actor` = case_when(
      `Actor 1 Age` > `Actor 2 Age` ~ `Actor 1 Gender`,
      `Actor 1 Age` < `Actor 2 Age` ~ `Actor 2 Gender`,
      TRUE ~ "Same age"
    ),
    `Avg Actor Age` = rowMeans(select(., `Actor 1 Age`, `Actor 2 Age`), na.rm = TRUE)
  )

joined_data <- joined_data %>%
  mutate(`Profit (Millions)` = round(profit / 1000000))

joined_data <- joined_data %>%
  mutate(`Age Group` = cut(`Avg Actor Age`, breaks = c(15, 30, 40, 50, 60, 100),
                           labels = c("15–30", "31–40", "41–50", "51–60", "60+"),
                           right = FALSE))

```

6.  Plot 1: Average Actor Age vs Profit

```{r}

p1 <- ggplot(joined_data, aes(x = `Age Group`, y = `Profit (Millions)`, fill = `Age Group`)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.1, color = "black", alpha = 0.8) +
  labs(title = "Profit Distribution by Actor Age Group",
       x = "Average Actor Age Group", y = "Profit (Millions USD)") +
  theme_minimal()

```

```{r}

# Profit vs Age Difference
p2 <- ggplot(joined_data, aes(x = `Age Difference`, y = `Profit (Millions)`)) +
  geom_point(alpha = 0.5, color = "#6a0dad") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Profit vs Age Gap",
    x = "Age Gap (Years)",
    y = "Profit (Millions USD)"
  ) +
  theme_minimal()

# ROI vs Age Difference
p3 <- ggplot(joined_data, aes(x = `Age Difference`, y = roi)) +
  geom_point(alpha = 0.5, color = "#ff5733") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "ROI vs Age Gap",
    x = "Age Gap (Years)",
    y = "Return on Investment (ROI)"
  ) +
  theme_minimal()

# Combine side by side
age_diff <- p2 + p3


```

```{r}

p4 <- ggplot(joined_data, aes(x = `Older Actor`, y = `Profit (Millions)`, fill = `Older Actor`)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Profit by Older Actor Gender",
    x = "",
    y = "Profit (Millions USD)"
  ) +
  theme_minimal()

# ROI by Older Actor
p5 <- ggplot(joined_data, aes(x = `Older Actor`, y = roi, fill = `Older Actor`)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "ROI by Older Actor Gender",
    x = "",
    y = "Return on Investment (ROI)"
  ) +
  theme_minimal()

older_actor_plot <- p4 + p5
```

### Remember

Please mention any additional files that you want the markers to review, eg. read me, meta data etc.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Motivation

One thing that’s always stood out to me in movies is how often we see older male leads paired with much younger female love interests. It’s a pattern that’s been normalized in Hollywood for decades — but I started wondering, does it actually pay off? Are these casting choices boosting a film’s profits, or are they just outdated traditions that filmmakers keep following without question?

### Data

To explore the link between casting age gaps and movie profitability, I used two main datasets.

The first came from [hollywoodagegap.com](hollywoodagegap.com), which lists romantic lead pairings in movies — including how old both actors were at the time of release, who was older, and the size of the age gap.

The second dataset, sourced from [GitHub Repository](https://github.com/ntdoris/movie-revenue-analysis), contained movie revenue details such as production budget, domestic and international gross, total profit, and return on investment (ROI).

After cleaning both datasets, I merged them using the movie title and release year. To make the analysis more meaningful, I only kept rows that matched accurately. I also created a new column to indicate whether the man, woman, or neither was older in each pairing, and calculated the average age of the two leads. This average actor age was then grouped into age brackets (15–30, 31–40, etc.) to help uncover any trends across different age groups.

### Analysis

#### Do younger romantic leads lead to higher profits?

```{r echo=FALSE}

p1

```


The first plot shows how profit varies with the average age of romantic leads, grouped into age bins (15–30, 31–40, etc.) using a violin plot.

Movies with younger leads (15–30) had the widest range of profits, including several major outliers. As the age increased, profits tended to drop and became more tightly clustered. The 60+ group was the most consistent but earned the least.

This suggests that younger lead pairings may be more profitable, likely due to broader audience appeal, stronger marketing, or more commercial genres.

#### What happens when the age gap between leads gets larger?

```{r echo=FALSE}

age_diff

```

This next set of plots looked at how the age gap between romantic leads relates to profit and ROI.

Both showed a clear downward trend — as the age gap increased, financial performance dropped. Most of the high-profit films had age gaps under 10 years. While a few exceptions exist, they were rare.

This suggests that larger age gaps may hurt a film’s success, possibly because audiences don’t relate as well to those pairings.


#### Does it matter whether the man or woman is older?

```{r echo=FALSE}

older_actor_plot

```

Here we compared profit and ROI based on which actor was older — the man, the woman, or if they were the same age.

Profit stayed fairly consistent across all groups. But when it came to ROI, same-age pairings performed best. Interestingly, films with older female leads did just as well — if not slightly better — than those with older male leads.

This challenges the long-standing stereotype that older women hurt box office success. The data suggests otherwise.

### Conclusion

This analysis suggests that some of Hollywood’s long-standing casting habits don’t hold up financially. Younger romantic leads do tend to generate more profit, but large age gaps between them can actually lower a film’s ROI. Interestingly, same-age pairings came out strongest, especially when it comes to return on investment.

There’s no real evidence that casting older men with younger women boosts box office performance. If anything, more balanced casting choices may be both creatively richer and commercially smarter. It might be time for studios to rethink what audiences really connect with — and what actually sells.

### References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
